---
title: "ggplot2 - Chapter 3 Examples and Excersizes"
output:
  html_document: 
    toc: true
    theme: united
  html_notebook: default
  pdf_document: 
    toc: true
    theme: united
---

This is a notebook of the examples and excersizes from "ggplot2 Elegant Graphics for Data Analysis"

```{r}
library(tidyverse)
library(gridExtra)
```
# Chapter 3

Excersizes and examples from Chapter 3.

Basich carts.

I used the [gridextra](https://cran.r-project.org/web/packages/gridExtra/index.html) package to plot multiple charts in a single window.

## Example from P51

```{r}
data.frame(
  x=c(3, 1, 5),
  y=c(2, 4, 6),
  label=c('a', 'b', 'c')) -> df

ggplot(df,
       aes(x, y, label=label)) +
  labs(x=NULL, y=NULL) + # Hides x and y axis
  theme(plot.title=element_text(size=12)) -> p  # Shrink plot title
```
```{r}
gridExtra::grid.arrange(p+geom_point() + ggtitle("point"), 
                        p + geom_text() + ggtitle("text"),
                        p + geom_bar(stat='identity') + ggtitle('bar'),
                        p + geom_tile() + ggtitle('raster'),
                        ncol=2)
```
```{r}
gridExtra::grid.arrange(p + geom_line() + ggtitle('line'),
                        p + geom_area() + ggtitle('area'),
                        p + geom_path() + ggtitle('path'),
                        p + geom_polygon() + ggtitle('polygon'),
                        ncol=2)
```
```{r}
gridExtra::grid.arrange(p + geom_dotplot() + ggtitle('dot'),
                        p + geom_rug() + ggtitle('rug'),
                        ncol=2)
```

## 3.3 Labels

Adding text to the plot.

Fonts are handled differently by different platforms and output types.  there are 2 packages which attempt to address this:

* [showtext](https://github.com/yixuan/showtext) by Yixuan Qiu
* [extrafont](htts://github.com/wch/extrafont) by Winston Chang

```{r}
data.frame(x=1, 
           y=3:1,
           family=c('sans', 'serif', 'mono')) -> df
ggplot(df, 
       aes(x=x, y=y)) +
  geom_text(aes(label=family,
                family=family)) -> p
```

Print the plot

```{r}
print(p)
```

## face example

```{r}
data.frame(x=1,
           y=3:1,
           face=c('plain', 'bold', 'italic')) -> df
ggplot(df, 
       aes(x=x, y=y)) +
  geom_text(aes(label=face,
                fontface=face))
```

## Alignment

```{r}
data.frame(
  x=c(1,1,2,2,1.5),
  y=c(1,2,1,2,1.5),
  text=c('bottom-left', 'bottom-right',
         'top-left', 'top-right', 'center')) -> df

ggplot(df,
       aes(x, y)) +
  geom_text(aes(label=text)) -> p1

ggplot(df,
       aes(x, y)) +
  geom_text(aes(label=text),
            vjust='inward',
            hjust='inward') -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

## size

ggplot2 uses mm for sizes.

There are 72.27 pts / inch so to convert from pts to mm multiply by (72.27 / 25.4).

plot the points, create a text label for each from the resp variable.  The second plot "nudges" down the text bo 0.25 mm.

The option to geom_text() **check_overlap** will check and remove overlapping text.  It simply uses only the first text that fills a space and does not display later text.  this can be handy when there are a lot of text values to clear things up a bit, but you can loose a lot of text.

```{r}
data.frame(
  trt=c('a', 'b', 'c'),
  resp=c(1.2, 3.4, 2.5)) -> df

ggplot(df,
       aes(x=resp, y=trt)) +
  geom_point() +
  geom_text(aes(label=paste0('(', resp, ')'))) + 
  xlim(1, 3.6) -> p1

ggplot(df,
       aes(x=resp, y=trt)) +
  geom_point() +
  geom_text(aes(label=paste0('(', resp, ')')),
            nudge_y=-0.25) + 
  xlim(1, 3.6) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

## geom_lable()

geom_label() is just like geom_text() but draws a rounded rectangle around the text with a sold background.  This can be handy when the background is busy, which, if you are following Tufte's guideliens, it will not be.

```{r}
data.frame(
  waiting=c(55, 80),
  eruptions=c(2,4.3),
  label=c('peak one', 'peak two')) -> label

ggplot(faithfuld,
       aes(waiting, eruptions)) +
  geom_tile(aes(fill=density)) +
  geom_label(data=label,
             aes(label=label)) -> p1

p1
```

## directlabels

[directlabels](https://github.com/tdhock/directlabels) is a package to plot labels directly on a plot "near" the data points they represent.

For more see the directlabels website http://directlabels.r-forge.r-projgect.org.

```{r}
ggplot(mpg,
       aes(x=displ, y=hwy, colour=class)) -> p

p + geom_point() -> p1
p + geom_point(show.legend=FALSE) +
  directlabels::geom_dl(aes(label=class),
                        method='smart.grid') -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)

```

## 3.4 Annotations

Annotations can be text, lines, colors, etc.  Anything to add additional meta data to a chart.

The folowing example shows unemployment data in use and adds information about what presedent was in office at the time.

```{r}
subset(presidential, start > economics$date[1]) -> pres

ggplot(economics) +
  geom_rect(
    aes(xmin=start,
        xmax=end,
        fill=party),
    ymin=-Inf, ymax=Inf, alpha=0.2,
    data=pres) +
  geom_line(aes(x=date, y=unemploy)) +
  scale_fill_manual(values=c('blue', 'red')) -> p1

p1 + geom_vline(
  aes(xintercept=as.numeric(start)),
  data=pres,
  colour='grey50',
  alpha=0.5) -> p2

p2 + geom_text(
  aes(x=start, y=2500, label=name),
  data=pres,
  size=3, vjust=0, hjust=0, nudge_x=50) -> p3

p2 + geom_text(
  aes(x=start, y=0, label=name),
  data=pres,
  size=3, vjust=0, hjust=0, nudge_x=50) -> p4

gridExtra::grid.table(pres)
gridExtra::grid.arrange(p1, p2, p3, p4, ncol=1)
```

# 3.5 Collective geoms

```{r}
data(Oxboys, package='nlme')
head(Oxboys)
```

Use **group** to group varaibles by a factor.  If there are multiple factors the data is grouped by, then include all the factors in the **interaction()** function, i.e. `aes(x=x, y=y, group=interaction(f1, f2))`.

```{r}
ggplot(Oxboys,
       aes(x=age, y=height, group=Subject)) +
  geom_point() +
  geom_line()
```

## Different groups on different layers

The **geom_smooth()** is applied to the data in the ggplot, since the grouping is applied at that level in the first chart, each group gets a smooth line.  The second chart, the grouping takes place at the **geom_line()** layer so the smoothing is applied to the complete ungrouped dataset and so is what we want.

```{r}
ggplot(Oxboys,
       aes(x=age,
           y=height,
           group=Subject)) +
  geom_line() +
  geom_smooth(method='lm', se=FALSE) -> p1

ggplot(Oxboys,
       aes(x=age,
           y=height)) +
  geom_line(aes(group=Subject)) +
  geom_smooth(method='lm', se=FALSE) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```


## Overriding default grouping

Exmaple of boxplot, which has default grouping along the x axis.

```{r}
ggplot(Oxboys,
       aes(x=Occasion, y=height)) +
  geom_boxplot() -> p1

p1
```

Let's see what it looks like to trace each boy through the data.  So add a **geom_line()**

```{r}
p1 + geom_line(colour='blue', 
               alpha=0.5)
```

Not what we wanted.  Since the default grouping is by the x values, the lines simply run vertically through the y values.  So we need to override the default grouping.

```{r}
p1 + geom_line(aes(group=Subject),
               colour='blue',
               alpha=0.5)


```

## Excersizes

1. Draw a boxplot of hwy for each value of cyl, without turning cyl into a factor.  What extra aesthetic to you need to set?

>> group=cyl is required to get the desired affect.

```{r}
head(mpg)
```
```{r}
ggplot(mpg,
       aes(x=cyl, y=hwy, group=cyl)) +
  geom_boxplot()
```

2. Modify the following so you get 1 boxplot per integer value of displ.

```
ggplot(mpg,
aes(x=displ, y=cty)) +
geom_boxplot()
```

```{r}
ggplot(mpg,
       aes(x=displ, y=cty, group=displ)) +
  geom_boxplot()
```

3. Why, from the example on discrete colors on a line, did you need to include the **aes(group=1)** for the **geom_line()**?  What happens if it's omitted? What is the difference between **aes(group=1)** and **aes(group=2)**.

Because there is only 1 point per grouping for the lines there is no way to know what color the line segments should be.

Adding the *group=1* provides a default grouping of all the values.  It can be any value, 1, 2, 3, 'spoon'.

```{r}
data.frame(x=1:3,
           y=1:3,
           colour=c(1,3,5)) -> df

ggplot(df,
       aes(x=x,
           y=y,
           colour=factor(colour))) + 
  geom_line(size=2) +
  geom_point(size=5) -> p1

ggplot(df,
       aes(x=x,
           y=y,
           colour=factor(colour))) + 
  geom_line(aes(group=1),
            size=2) +
  geom_point(size=5) -> p2

ggplot(df,
       aes(x=x,
           y=y,
           colour=factor(colour))) + 
  geom_line(aes(group=2),
            size=2) +
  geom_point(size=5) -> p3


ggplot(df,
       aes(x=x,
           y=y,
           colour=factor(colour))) + 
  geom_line(aes(group='spoon'),
            size=2) +
  geom_point(size=5) -> p4

gridExtra::grid.arrange(p1, p2, p3, p4, ncol=2)
```

4. How many bars per graph?

The first, 3.  The second, 1 for each distinct hwd value.  For the thrid 1 for each observation.

They are stacked.

```{r}
unique(mpg$drv)
```
```{r}
ggplot(mpg,
       aes(x=drv)) + 
  geom_bar() -> p1

ggplot(mpg,
       aes(x=drv, fill=hwy, group=hwy, colour='white')) +
  geom_bar() -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```
```{r}
library(dplyr)
mpg %>%
  arrange(hwy) %>%
  mutate(id=seq_along(hwy)) -> mpg2

head(mpg2)
```
```{r}
ggplot(mpg2,
       aes(x=drv,
           fill=hwy,
           group=id,
           colour='white')) + 
  geom_bar()
```

Install the babynames package.

5. Fix the graph, why is Hadley unhappy?

Hadley is becomming a womans name.

```{r}
library(babynames)

head(babynames)
```
```{r}
dplyr::filter(babynames,
              name=="Hadley") -> hadley

hadley
```
```{r}
ggplot(hadley,
       aes(x=year, y=n)) +
  geom_line() -> p1

ggplot(hadley,
       aes(x=year, y=n, colour=sex)) +
  geom_line() -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

## 3.6 Surface Plots

ggplot2 does not directly support 3d plots but does provide ways of representing 3d data such as contours, colored tiles and buble plots.

For interactve 3d plots see RGL, http://rgl.neoscientists.org/about.shtml.

```{r}
head(faithfuld)
```

..level.. used as the factor for the _colour_ aesthetic is a calculated value produced by the geom_contour statistic.

```{r}
ggplot(faithfuld,
       aes(x=eruptions, y=waiting)) +
  geom_contour(aes(z=density, colour=..level..)) -> p1

ggplot(faithfuld,
       aes(x=eruptions, y=waiting)) +
  geom_raster(aes(fill=density)) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

Buble plots work best with smaller datasets.

```{r}
faithfuld[seq(1, nrow(faithfuld), by=10),] -> small

ggplot(small,
       aes(x=eruptions, y=waiting)) +
  geom_point(aes(size=density), 
             alpha=1/3) + 
  scale_size_area()
```

## 3.7 Maps
### Vector boundaries

Vectory boundaries are represented by a data.frame with 1 corner / row and 4 variables / row.

latitude, longitude, group and id.  Group is the overall group, such as Hawaii and id is a distinct area
with it, such as each individual island.

**NOTE:** Install the maps package: `install.packages('maps')` for this to work.

```{r}
map_data('county', 'michigan') %>% 
  select(lon=long, lat, group, id=subregion) -> mi_counties

head(mi_counties)
```

The **coord_quickmap()** is used to adjust the aspect of the data to take into account the earth is round and degrees of lat closer as one approaches the poles.  Also see the more accurate but compute intensive **coord_map()**.

```{r}
ggplot(mi_counties,
       aes(x=lon, y=lat)) +
  geom_polygon(aes(group=group)) +
  coord_quickmap() -> p1

ggplot(mi_counties,
       aes(x=lon, y=lat)) +
  geom_polygon(aes(group=group),
               fill=NA,
               colour='grey50') +
  coord_quickmap() -> p2

ggplot(mi_counties,
       aes(x=lon, y=lat)) +
  geom_polygon(aes(group=group)) -> p3

ggplot(mi_counties,
       aes(x=lon, y=lat)) +
  geom_polygon(aes(group=group),
               fill=NA,
               colour='grey50') -> p4

gridExtra::grid.arrange(p1, p2, p3, p4, ncol=2)
```

Much more accurate map data for the U.S. including historic data is available from The USAboudnaries package; https://github.com/ropensci/USAboundaries.  `install.packages('USAboundaries')`

Here is an example of getting at the data and displaying the U.S. as of 1840.

See also the tigris package https://github.com/ walkerke/tigris package.  `install.package('tigris')` and http://naturalearthdata.com/, https://cran.r-project.org/package=osmar.

```{r}
library(USAboundaries)

us_states('1840-03-12') -> states_1840

data.frame() -> df
for ( j in 1:length(states_1840@polygons) ) {
  for ( i in states_1840@polygons[[j]]@plotOrder ) {
    rbind(df, data.frame(lon=states_1840@polygons[[j]]@Polygons[[i]]@coords[,1],
                         lat=states_1840@polygons[[j]]@Polygons[[i]]@coords[,2],
                         group=as.numeric(states_1840@polygons[[j]]@ID),
                         id=i)) -> df
  }
}

ggplot(data=df,
       aes(x=lon, y=lat)) +
  geom_polygon(aes(group=interaction(id, group)),
               fill=NA,
               colour='grey50') +
  coord_quickmap() -> p

p
```

And same map but leveraging the sp library.

```{r}
fortify(states_1840) -> states_1840df

head(states_1840df)
```
```{r}
ggplot(data=states_1840df,
       aes(x=long, lat)) +
  geom_polygon(aes(group=group), colour='grey50', fill=NA) + 
  coord_quickmap() -> p

p
```

Shape (.shp) files may be read directly with the maptools::readShapSpacial() function.


```{r}
download.file('http://www2.census.gov/geo/tiger/GENZ2015/shp/cb_2015_us_nation_5m.zip',
              'cb_2015_us_nation_5m.zip')
```

Unziped file manually

```{r}
library(maptools)
readShapeSpatial('cb_2015_us_nation_5m.shp') -> ussp
```

## Point data

City locations are "point data".  These are almost always combined with a background map or they are just points on the screen with no reference.

```{r}
maps::us.cities %>%
  tbl_df() %>%
  filter(country.etc == "MI") %>%
  select(-country.etc, lon=long) %>%
  arrange(desc(pop)) -> mi_cities

mi_cities
```

Graph just the points

```{r}
ggplot(mi_cities,
       aes(x=lon, y=lat)) +
  geom_point(aes(size=pop)) +
  scale_size_area() +
  coord_quickmap() -> p1

p1
```

Now let's do it again but draw the state as well.

```{r}
ggplot(mi_cities,
       aes(x=lon, y=lat)) +
  geom_polygon(aes(x=lon, y=lat, group=group),
               data=mi_counties,
               fill=NA,
               colour='grey50') +
  geom_point(aes(size=pop), 
             colour='red',
             alpha=0.5) +
  scale_size_area() + 
  coord_quickmap() -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

## 3.7.3 Raster images

Getting raster images can be slow, cache them if you get them!!!!

```{r}
library(ggmap)

if ( file.exists('mi_raster.rds') ) {
  readRDS('mi_raster.rds') -> mi_raster
} else {
  c(min(mi_counties$lon), min(mi_counties$lat),
    max(mi_counties$lon), max(mi_counties$lat)) -> bbox
  ggmap::get_openstreetmap(bbox,
                           scale=8735660) -> mi_raster
  saveRDS(mi_raster, 'mi_raster.rds')
}
```

```{r}
ggmap::ggmap(mi_raster)
```

```{r}
ggmap::ggmap(mi_raster) +
  geom_point(aes(size=pop),
             data=mi_cities,
             colour='red',
             alpha=0.5) + 
  scale_size_area()
```

## 3.7.4 Area metadata


```{r}
tbl_df(midwest) %>%
  filter(state == 'MI') %>%
  mutate(county=tolower(county)) %>%
  select(county, area, poptotal, percwhite, percblack) -> mi_census

mi_census
```

Merge this with the mi_counties data

```{r}
left_join(mi_census, mi_counties, by=c('county' = 'id')) -> census_counties
census_counties
```

```{r}
ggplot(census_counties,
       aes(x=lon, y=lat, group=county)) +
  geom_polygon(aes(fill=poptotal)) +
  coord_quickmap() -> p1

ggplot(census_counties,
       aes(x=lon, y=lat, group=county)) +
  geom_polygon(aes(fill=percwhite)) +
  coord_quickmap() -> p2

ggplot(mi_counties,
       aes(x=lon, y=lat)) +
  geom_polygon(aes(group=group),
               fill=NA,
               colour='grey50') +
  coord_quickmap()-> p3

gridExtra::grid.arrange(p1, p2, p3, ncol=2)
```

## 3.8 Revealing uncertainty

4 geoms available

* Discrete x, range: geom_errorbar(), geom_linerange()
* Discrete x, range & cener: geom_crossbar(), geom_pointrange()
* Continuous x, range: geom_ribbon()
* Continuous x, range & center: geom_smoth(stat='identiry')

```{r}
c(18,11,16) -> y
data.frame(x=1:3,
           y=y,
           se=c(1.2, 0.5, 1.0)) -> df

ggplot(df,
       aes(x=x, y=y, ymin=y-se, ymax=y+se)) +
  geom_line(colour='red') -> base

base + geom_crossbar() -> p1
base + geom_pointrange() -> p2
base + geom_smooth(stat='identity') -> p3

gridExtra::grid.arrange(p1, p2, p3, ncol=3)

```

```{r}
base + geom_errorbar() -> p1
base + geom_linerange() -> p2
base + geom_ribbon(alpha=0.3, colour='grey50') -> p3

gridExtra::grid.arrange(p1, p2, p3, ncol=3)
```

## 3.9 Weighted data

For lines and points we can use the size aesthetic.

```{r}
ggplot(midwest,
       aes(x=percwhite, y=percbelowpoverty)) + 
  geom_point() -> p1

ggplot(midwest,
       aes(x=percwhite, y=percbelowpoverty)) +
  geom_point(aes(size=poptotal / 1e6)) +
  scale_size_area("Population\n(millions)",
                 breaks=c(0.5, 1, 2, 4)) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

Weighting can be passed to summary functions with the weight= parm.  This is not seen in the data directly but affects the smoothing results.

```{r}
ggplot(midwest,
        aes(x=percwhite, y=percbelowpoverty)) + 
  geom_point() + 
  geom_smooth(method=lm, size=1) -> p1

ggplot(midwest,
       aes(x=percwhite, y=percbelowpoverty)) +
  geom_point(aes(size=poptotal/1e6)) +
  geom_smooth(aes(weight=poptotal), 
              method=lm,
              size=1) +
  scale_size_area(guide='none') -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

Example of the affects of weighting on a histogram.  Note that the weighting changes what is counted  It goes from Counties to Population.

```{r}
ggplot(midwest,
       aes(x=percbelowpoverty)) + 
  geom_histogram(binwidt=1) +
  ylab("Counties") -> p1

ggplot(midwest,
       aes(x=percbelowpoverty)) +
  geom_histogram(aes(weight=poptotal),
                 binwidth=1) +
  ylab('Population (1000)s') -> p2

grid.arrange(p1, p2, ncol=2)
```

## 3.10 Diamonds data

Use binwidth to find the best view of the data.
Use things like xlim() to limit the x axis range to focus in on a key area.

```{r}
ggplot(diamonds, 
       aes(depth)) +
  geom_histogram() -> p1

ggplot(diamonds,
       aes(depth)) +
  geom_histogram(binwidth=0.1) +
  xlim(55, 70) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

Setting position='fill' for the geom_histogram causes the use of the density value instead of straight count.  This is the number of items at that point / total count of itmes * binwidth.

```{r}
ggplot(diamonds,
       aes(depth)) +
  geom_freqpoly(aes(colour=cut),
                binwidth=0.1,
                na.rm=TRUE) +
  xlim(58, 68) +
  theme(legend.position='none') -> p1

ggplot(diamonds,
       aes(depth)) +
  geom_histogram(aes(fill=cut),
                 binwidth=0.1,
                 position='fill',
                 na.rm=TRUE) +
  xlim(58, 68) +
  theme(legend.position='none') -> p2

ggplot(diamonds,
       aes(depth)) +
  geom_histogram(aes(fill=cut),
                 binwidth=0.1,
                 na.rm=TRUE) +
  xlim(58, 68) +
  theme(legend.position='none') -> p3

gridExtra::grid.arrange(p1, p2, p3, ncol=2)
```

```{r}
ggplot(diamonds,
       aes(x=depth)) +
  geom_density(na.rm=TRUE) +
  xlim(58, 68) + 
  theme(legend.position='none') -> p1

ggplot(diamonds,
       aes(x=depth,
           fill=cut,
           colour=cut)) +
  geom_density(alpha=0.2,
               na.rm=TRUE) +
  xlim(58, 68) +
  theme(legend.position='none') -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

## boxplot

```{r}
ggplot(diamonds,
       aes(clarity, depth)) +
  geom_boxplot() -> p1

ggplot(diamonds,
       aes(carat, depth)) +
  geom_boxplot(aes(group=cut_width(carat, 0.1))) +
  xlim(NA, 2.05) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

## violin plot

```{r}
ggplot(diamonds,
       aes(clarity, depth)) +
  geom_violin() -> p1

ggplot(diamonds,
       aes(carat, depth)) +
  geom_violin(aes(group=cut_width(carat, 0.1))) + 
  xlim(NA, 2.05) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

## Excersizes

* What binwidth tells most ineresting story about distribution of carat?

0.01 - Shows a repeated downward slope between whole values.

```{r}
ggplot(diamonds,
       aes(x=carat)) + 
  geom_histogram() -> p1

ggplot(diamonds,
       aes(x=carat)) +
  geom_histogram(binwidth=0.1) -> p2

ggplot(diamonds,
       aes(x=carat)) +
  geom_histogram(binwidth=0.01) -> p3

gridExtra::grid.arrange(p1, p2, p3, ncol=2)
```

### Draw a historgram of price.
### What interesting patters do you see?  There is a bump in the data between 4-5K.

```{r}
ggplot(diamonds,
       aes(x=price)) +
  geom_histogram() -> p1

ggplot(diamonds,
       aes(x=price)) +
  geom_histogram(binwidth=100) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

### How does distribution of price vary with clarity?

```{r}
ggplot(diamonds,
       aes(x=price, colour=clarity, fill=clarity)) +
  geom_density(alpha=0.2)  -> p1


gridExtra::grid.arrange(p1, ncol=2)
```

## Overplotting

When using a scatter plot points may fall on each other obscuring the actual data making it hard to determine anything from the data.  Some approaches to address this are:

Change size or type of point to show data.

```{r}
data.frame(x=rnorm(2000),
           y=rnorm(2000)) -> df

ggplot(df,
       aes(x=x, y=y)) +
  xlab(NULL) + ylab(NULL) -> norm

gridExtra::grid.arrange(norm + geom_point(),
                        norm + geom_point(shape=1),
                        norm + geom_point(shape='.'),
                        ncol=3)
```

Change the gransparancy of the points so that overlapping will show darker.

```{r}
gridExtra::grid.arrange(norm + geom_point(alpha=1/3),
                        norm + geom_point(alpha=1/5),
                        norm + geom_point(alpha=1/10),
                        ncol=3)
```

Another approach is to 'bin' points in certain arias and plot the density with colour.

```{r}
gridExtra::grid.arrange(norm + geom_bin2d(),
                        norm + geom_bin2d(bins=10),
                        ncol=2)
```

Squares from the geom_bin2d aren't very visually appealing, using geom_hex from the hexbin package makes nicer images

```{r}
library(hexbin)

gridExtra::grid.arrange(norm + geom_hex(),
                        norm + geom_hex(bins=10),
                        ncol=2)
```


## Statistical Summaries

```{r}
ggplot(diamonds,
       aes(color)) + 
  geom_bar() -> p1

ggplot(diamonds,
       aes(color, price)) +
  geom_bar(stat='summary_bin',
           fun.y=mean) -> p2

gridExtra::grid.arrange(p1, p2, ncol=2)
```

```{r}
ggplot(diamonds,
       aes(x=table, y=depth)) +
  geom_bin2d(binwidth=1, na.rm=TRUE) +
  xlim(50, 70) + 
  ylim(50, 70) -> p1

ggplot(diamonds,
       aes(x=table, y=depth, z=price)) +
  geom_raster(binwidth=1, 
              stat='summary_2d', 
              fun=mean,
              na.rm=TRUE) +
  xlim(50, 70) + 
  ylim(50, 70) -> p2

p1
p2
```



