---
title: "ggplot2 - Chapter 4 Examples and Excersizes"
output:
  pdf_document:
    toc: yes
  html_notebook: default
  html_document:
    theme: united
    toc: yes
---

his is a notebook of the examples and excersizes from "ggplot2 Elegant Graphics for Data Analysis"

```{r}
library(tidyverse)
library(gridExtra)
```

# Chapter 4

## Building a scatterplot
### Mapping aesthetics to data

**aesthetics** are the attributes which describe a chart.

See the `vignett("ggplot2-specs")` for colory, size, line-type and font values.

In the example below the aesthetics applied are to set _displ_ to the _x axis_, _hwy_ to the _y axis_ and set a different color for each _cyl_ value as a factor.  The result is a dataset that looks like:


```{r}
select(mpg, x=displ, y=hwy, colour=cyl) %>%
  head()
```

This new dataset can then be drawn as any chart type, although some may not make sense.

```{r}
ggplot(mpg,
       aes(x=displ, y=hwy, colour=factor(cyl))) +
  theme(legend.position='none') -> p

gridExtra::grid.arrange(p + geom_point(),
                        p + geom_line(),
                        p + geom_bar(stat='identity', position='identity', fill=NA),
                        ncol=3)
```

Graphs are made up of _geometric shapes_ or _geoms_.  Some special sinlge geom charts exist like scatterplot, bubble chart, line chart, etc.  Other more complex don't have single names and are described individually.

An example is a scatter plot of data with a per group regression line through each group.

```{r}
ggplot(mpg,
       aes(x=displ, y=hwy, colour=factor(cyl))) + 
  geom_point() +
  geom_smooth(method='lm')
```

### Scaling

The x and y values have to be "scaled" so R and the graphics enging (grid) can plot them.  By default the values are mapped to a value between [0,1].  The underlying **coord** system takes care of mapping that to the appropriate pixel on the output device.

Colors are mapped to a 6 hex digit value that cooresponds to a screen color.

For the example above the dataset to map would include the default size (1) and shape (19 - filled ciricle) and default assigned color.  Soemthing like:

```{r}
ds = dplyr::select(mpg, x=displ, y=hwy, colour=cyl)
ds$size = 1
ds$shape = 19

ds[ds$colour == 4, 'colour'] = '#F12345'
ds[ds$colour == 5, 'colour'] = '#F23456'
ds[ds$colour == 6, 'colour'] = '#F34567'
ds[ds$colour == 8, 'colour'] = '#F45678'

head(ds)
```

Now for a more complicated example

```{r}
ggplot(mpg,
       aes(x=displ, y=hwy)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~year)
```

This is an example with multiple layers (points and smooth lines), statistics (the smooth line).

### Scaling happens in 3 parts

* Transform
* Train
* Map

Scale transformations occur before statistical transformations so stats are calculated on transformed data.

## components of layered graphics

1. Default dataset and aesthetic mappings
2. One or more layers composed of a geom, a statistical transformation, position adjustment and optionally a dataset and aesthetic mappings.
3. One scale for each aesthetic mapping.
4. A coordinate system.
5. The facetting specification

### Layers

5 components to layers

1. Data
2. Aesthetic mappings
3. A statistical transformation (stat)
4. A geom
5. A postition adjustment

### Scales

A scale is requried for each aesthetic used on a plot.  This controlls mapping form aesthtic to data.

### Coordinate System

Controls how points are mapped to the output device, usually by (x, y) pairs but there could be more (ggplot does not support more than 2 dimentions)  There are 2 commonly used coordinate systems, Cartesian and Polar. 

The coordinate system can modify the look of the geom.  i.e. a bar in a bar chart looks like a segment of a circle in polar coordinates.

### Facetting

Break out into multiple charts by some factor(s)


