---
title: "ggplot2 - Chapter 7 Examples and Excersizes"
output:
  pdf_document:
    toc: yes
  html_notebook: default
  html_document:
    theme: united
    toc: yes
  word_document:
    toc: yes
---

his is a notebook of the examples and excersizes from "ggplot2 Elegant Graphics for Data Analysis"

```{r}
library(tidyverse)
library(gridExtra)
```

# Positioning

## faceting

3 faceting functions

* facet_null() - defalt 1 chart
* facet_wrap()
* facet_grid()

```{r}
subset(mpg,
       cyl != 5 & drv %in% c('4', 'f') & class != '2seater') -> mpg2

ggplot(mpg2,
       aes(x=displ, y=hwy)) +
  geom_blank() + 
  labs(x=NULL, y=NULL) -> pb

gridExtra::grid.arrange(pb + facet_wrap(~ class, ncol=3),
                        pb + facet_wrap(~ class, 
                                        ncol=3,
                                        as.table=FALSE),
                        ncol=2)
```

```{r}
gridExtra::grid.arrange(pb + facet_wrap(~ class, nrow=3),
                        pb + facet_wrap(~ class,
                                        nrow=3,
                                        dir='v'),
                        ncol=2)
```

### facet grid

. ~ <var> spreads items accross colums

```{r}
pb + facet_grid(. ~ cyl)
```

<var> ~ . spreads items down rows

```{r}
pb + facet_grid(drv ~ .)
```

<var> ~ <var> spreads accross columns and down rows

```{r}
pb + facet_grid(drv ~ cyl)
```

Multiple values can be used in the formula by "adding" them (+).

### Controlling scales

by default scales for charts are "fixed" at the values for the largest range.  There are 4 options, however.

* scales='fixed' (default)
* scales='free_x' - X scales are relative to data in that chart
* scales='free_y' - Y scales are relative to data in that chart
* scales='free' - Both X and Y are relatove to the data.

_facet_grid()_ has the constraint that all charts in a row have the same y scale and all charts in a column have the same x scale.

Fixed scales make it easier to see relations betwen charts.  Free scales make it easier to see details of each chaart.

```{r}
ggplot(mpg2,
       aes(x=cty, y=hwy)) + 
  geom_abline() +
  geom_jitter(width=0.1, height=0.1) -> p

gridExtra::grid.arrange(p + facet_wrap(~cyl),
                        p + facet_wrap(~cyl, scales='free'),
                        nrow=2)
```

free scales are also good for time series with different scales.  First, make data long.

```{r}
economics_long
```

```{r}
ggplot(economics_long,
       aes(x=date, y=value)) + 
  geom_line() + 
  facet_wrap(~variable, scales='free_y', ncol=1)
```

_facet_grid()_ has an additional parameter, _space_, which has same values as scales and controls how much of the available space each chart will get.

```{r}
reorder(mpg2$model, mpg2$cty) -> mpg2$model
reorder(mpg2$manufacturer, -mpg2$cty) -> mpg2$manufacturer

ggplot(mpg2, 
       aes(x=cty, y=model)) +
  geom_point() + 
  facet_grid(manufacturer ~ ., 
             scales='free',
             space='free') + 
  theme(strip.text.y = element_text(angle=0))
```

### Missing faceting variables

It displays in every facet.

```{r}
data_frame(x=1:3, y=1:3, gender=c('f', 'f', 'm')) -> df1
data_frame(x=2, y=2) -> df2

ggplot(df1,
       aes(x=x, y=y)) +
  geom_point(data=df2,
             colour='red',
             size=2) +
  geom_point() + 
  facet_wrap(~gender)
```

### Grouping vs. faceting

Faceting is a form of grouping.  Faceting has advantages in that each group is clear but disadvantage in seeing small differences betweengroups.

```{r}
data.frame(
  x=rnorm(120, c(0, 2, 4)),
  y=rnorm(120, c(1, 2, 1)),
  z = letters[1:3]) -> df

ggplot(df,
       aes(x=x, y=y)) +
  geom_point(aes(colour=z))
```

```{r}
ggplot(df,
       aes(x=x, y=y)) +
  geom_point() +
  facet_wrap(~z)
```

Comparisons betweenfacets can often be aided by thoughtful annotations such as adding the mean of each group to each facet.

```{r}
library(wesanderson)

group_by(df, z) %>%
  summarize(x=mean(x), y=mean(y)) %>%
  rename(z2=z) -> df_sum

ggplot(df,
       aes(x=x, y=y)) +
  geom_point() +
  geom_point(data=df_sum,
             aes(colour=z2),
             size=4) + 
  scale_colour_manual(name='z means',
                      values=wes_palette('GrandBudapest')) +
  facet_wrap(~z)
```

Another approach is to put all the data in the background in a subdued color.

```{r}
select(df, -z) -> df2

ggplot(df,
       aes(x=x, y=y)) +
  geom_point(data=df2, 
             aes(x=x, y=y)) +
  geom_point(aes(colour=z)) + 
  scale_colour_manual(name='z values',
                      values=wes_palette('Moonrise2')) +
  facet_wrap(~z)
```

### Continuous variables

To facet continuous variables they must first be "discretised".  There are 2 ggplot2 functions available for this

* cut_interval(x, n) - devides data into n bins of the same length (relative to data range).
* cut_width(x, width) - devices data into bins with width items in each.
* cut_number(x, n = 10) - devices data into n bins of about the same number of items.

```{r}
cut_width(mpg2$displ, 1) -> mpg2$disp_w
cut_interval(mpg2$displ, 6) -> mpg2$disp_i
cut_number(mpg2$displ, 6) -> mpg2$disp_n

mpg2
```

```{r fig.width=8, fig.height=3}
ggplot(mpg2,
       aes(x=cty, y=hwy)) +
  geom_point() +
  labs(x=NULL, y=NULL) -> plot

plot + facet_wrap(~disp_w, nrow=1)
```
```{r fig.width=8, fig.height=3}
plot + facet_wrap(~disp_i, nrow=1)
```
```{r fig.width=8, fig.height=3}
plot + facet_wrap(~disp_n, nrow=1)
```

## Coordinate systems

There are 2 types of coordinate systems, linear and non-linear.  Linear include:

* coord_cartesian()
* coord_flip()
* coord_fixed()

Non-linear include:

* coord_map()/coord_quickmap()
* coord_polar()
* coord_trans()

### Zooming into a plot with coord_cartesian

```{r}
ggplot(mpg,
       aes(x=displ, y=hwy)) +
  geom_point() +
  geom_smooth() -> base

gridExtra::grid.arrange(base, 
                        base + scale_x_continuous(limits=c(5,7)),
                        base + coord_cartesian(xlim=c(5,7)),
                        ncol=3)
```

### Flipping the axes with coord_flip()

```{r}
ggplot(mpg,
       aes(x=displ, y=cty)) +
  geom_point() +
  geom_smooth() -> p1

ggplot(mpg,
       aes(x=cty, y=displ)) +
  geom_point() +
  geom_smooth() -> p2

gridExtra::grid.arrange(p1, p2, p1 + coord_flip(), ncol=2)
```

## Non-linear coordinate systems

```{r}
data_frame(x=50, y=50) -> rect
data_frame(x=c(1,200), y=c(100,1)) -> line

ggplot(mapping=aes(x=x, y=y)) +
  geom_tile(data=rect, aes(width=50, height=50)) + 
  geom_line(data=line) +
  labs(x=NULL, y=NULL) -> base

gridExtra::grid.arrange(base, 
                        base+coord_polar('x'),
                        base+coord_polar('y'),
                        ncol=3)
```

```{r}
gridExtra::grid.arrange(base + coord_flip(),
                        base + coord_trans(y='log10'),
                        base + coord_fixed(),
                        ncol=3)
```

### Transformations with coord_trans()

scale transform happens before statistics are applied and do not affecdt the shape of the geom.  coord transforms occur after statistics are applied and do.  So you can apply a transform at the scale level then use coord_trans to backtransform it for interpretation.

```{r}
ggplot(diamonds,
       aes(carat, price)) +
  stat_bin2d() +
  geom_smooth(method='lm') + 
  labs(x=NULL, y=NULL) +
  theme(legend.position='none') -> base

base + 
  scale_x_log10() +
  scale_y_log10() -> base_log10

scales::exp_trans(10) -> pow10


gridExtra::grid.arrange(base,
                        base_log10,
                        base_log10 + coord_trans(x=pow10, y=pow10),
                        ncol=3)
```

### Polar coords with coord_polar()

Turning bar chart into pie or bullseye charts

```{r fig.width=8, fig.height=3}
ggplot(mtcars,
       aes(x=factor(1), fill=factor(cyl))) +
  geom_bar(width=1) +
  theme(legend.position='none') +
  scale_x_discrete(NULL, expand=c(0,0)) + 
  scale_y_continuous(NULL, expand=c(0,0)) -> base

gridExtra::grid.arrange(base + ggtitle('stacked bar'),
                        base + coord_polar(theta='y') + ggtitle('pieified bar chart'),
                        base + coord_polar() + ggtitle('bullseyeified bar chart'),
                        ncol=3)
```

Interesting affect when x is not a single value

```{r fig.width=8, fig.height=3}
ggplot(mtcars,
       aes(x=factor(carb), fill=factor(cyl))) +
  geom_bar(width=1) +
  theme(legend.position='none') +
  scale_x_discrete(NULL, expand=c(0,0)) + 
  scale_y_continuous(NULL, expand=c(0,0)) -> base

gridExtra::grid.arrange(base + ggtitle('stacked bar'),
                        base + coord_polar(theta='y') + ggtitle('pieified bar chart'),
                        base + coord_polar() + ggtitle('bullseyeified bar chart'),
                        ncol=3)
```

### Map projections with coord_map()

maps are intrisically printed on a sphere and must be adjusted.  There are 2 ways to do this with ggplot2.

```{r fig.width=8, fig.height=3}
ggplot(map_data('nz'),
       aes(x=long,
           y=lat,
           group=group)) +
  geom_polygon(fill='white',
               colour='black') + 
  labs(x=NULL, y=NULL) -> nzmap

gridExtra::grid.arrange(nzmap,
                        nzmap + coord_quickmap(),
                        ncol=2)
```

coord_map() uses the **mapproj** package to project the map.

```{r}
map_data('world') -> world
ggplot(world,
       aes(x=long,
           y=lat,
           group=group)) +
  geom_path() + 
  scale_y_continuous(name=NULL,
                     breaks=(-2:3) * 30, labels=NULL) +
  scale_x_continuous(name=NULL,
                     breaks=(-4:4) * 45, labels=NULL) -> world

gridExtra::grid.arrange(world, 
                        world+coord_map(),
                        world+coord_map('ortho'),
                        world+coord_map('stereographic'),
                        ncol=2)
```



